MEGA: New metrics to evaluate latent variable generative models

### 2021-08-28 Draft.tex
Pushing both MEGA files

\title{Assessing latent variables models by studying the moment estimator gap}

@@

\maketitle

@@

@@

We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature there is no good metric for such thing at the moment.

@@

\begin{align}
\Ex[X^2] = \Vx(X)+ (\Ex[X])^2 \approx \frac{\sum_{i=1}^n(X_i-\bar{X})^T(X_i-\bar{X})}{n-1} + \bar{X}^T\bar{X} = \frac{\sum_{i=1}^n X_i^TX_i}{n-1}
\label{lhs}
\end{align}

@@

\begin{align}
\Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
&\approx \frac{1}{m} \sum_{i=1}^m \left[\Vx(\x|\z=z_i) + \Ex(\x|\z=z_i)^T\Ex(\x|\z=z_i)\right]
\label{rhs}
\end{align}

@@

@@

Additional, both norms can be computed quite efficiently. To compute to Frobenius norm, we need square of the MEGA , $m^2$ operations, and then sum its diagonal which results in $m^2+m$ operations. The largest eigenvalue of a matrix can be estimated with the von Mises iteration \cite{mises29}, where each of the $p$ iterations will require $m^2$ operations resulting in $pm^2$ operations in total.

\subsection{Implementation of the selected norms}

\section{Experiments and results} \label{expe}

>>>>

\title{MEGA: New metrics to evaluate latent variable generative models}

@@

\maketitle
\begin{abstract}
In this article we introduce a new metric built to measure how latent variable models, such as Gaussian Mixture Models or Variational AutoEncoders, capture the first and second moments of data sets. We propose to compare the data moments to a moment estimator that uses both components of latent variable models. In order to make the gap between those two estimator digestible, we propose using the Frobenius norm of this gap to measure if the latent variable model did capture the moments of the data set. We discuss our metric behaviour under various scenarios and demonstrate it on multiple data sets with various latent variable models. Finally, we discuss various extensions: we discuss how to extend this metric to sequential latent variable model such as Hidden Markov Models and State-Space models, how to extend this metric to higher moments and the possibility of training model by minimizing this metric.
\bigskip

%\noindent

\textbf{Keywords} : Moment estimators, Latent Variable models, Frobenius norm, Gaussian Mixture Models, Variational Auto-Encoders
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\pagebreak

@@

The VAE is also a member of the LVGM family. It is assumed that $\z$ is a continuous variable where $p(\z)$ is assumed to be $N(0,I)$. $p(\x|\z)$ can be any parametric distribution where $f(\z)$ outputs the parameters of this distribution. For instance, if $p(\x|\z)$ is normal then $f(\z)$ will output a mean and a variance parameter, $f: \mathcal{R}_m \rightarrow \mathcal{R}_d \times \mathcal{R}_(d,d)^+$.

One novelty of VAEs is that the function $f$ proposed is much more flexible than a linear combination; it is a neural network.  In turns, this makes the posterior distribution $p(\z|\x)$ intractable and which prevent the model from being fitted by the EM algorithm. The solution proposed is to assume a variational distribution $q(\z|\x)$ and optimize the system by maximizing the Evidence Lower BOund (ELBO), a lower bound of the observed-data log-likelihood.

@@

We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature the lack of a reliable metric to asses latent variable models is obvious. At the moment, researchers rely either on the likelihood of the data under the fitted model or the heuristic analysis of generated data point. Both techniques have their fair share of problems.

\cite{Kingma13,

@@

\begin{align}
\text{DE} = \Ex[X^2] = \Vx(X)+ (\Ex[X])^2 \approx \frac{\sum_{i=1}^n(X_i-\bar{X})^T(X_i-\bar{X})}{n-1} + \bar{X}^T\bar{X} = \frac{\sum_{i=1}^n X_i^TX_i}{n-1}
\label{lhs}
\end{align}

@@

\begin{align}
\text{FME}= \Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
&\approx \frac{1}{m} \sum_{i=1}^m \left[\Vx(\x|\z=z_i) + \Ex(\x|\z=z_i)^T\Ex(\x|\z=z_i)\right]
\label{rhs}
\end{align}

@@

\subsection{Additional justification}

\subsection{More moments}

@@

Finally, the metrics we propose to evaluate the quality of the fit for the second moment is:

\begin{align}
\text{MEGA-F} &= |\text{DE-FME}|_F. \\
\text{MEGA-Op} &= |\text{DE-FME}|_{op}. \\
\label{MEGAs}
\end{align}

\subsection{Implementation of the selected norm}

Implementing the Frobenius norm is quite straight forward. To compute to Frobenius norm, we need square of the MEGA , $m^2$ operations, and then sum its diagonal which results in $m^2+m$ operations.

The largest eigenvalue of a matrix can be estimated with the von Mises iteration \cite{mises29}, where each of the $p$ iterations will require $m^2$ operations resulting in $pm^2$ operations in total.

\section{Experiments: Assessing VAE's ability to capture second order properties} \label{expe}
